A Simplified Formulation for Molecular Dynamics with Rigid Bodies

### 2016-07-06 rigid_bodies.tex
Changes in the first integration scheme for NVT

Analytical solutions exist for some simpler hypothetical cases related to Eq.~\ref{eq:EDO_system_alternative}. Let us analyze three of them, which will be useful later on to derive a complete approximate solution. In what follows, subscript $0$ denotes ``initial values''. We start by considering a hypothetical Hamiltonian $\mathcal{H}_t = (1/2m) \tr{\vt p} \vt p$, whose resulting equations of motion are $\dot{\vt r} = {\vt p}/m$, $\dot{\vt p} = 0$, $\dot{\vt q} = 0$, and $\dot{\vt \pi} = 0$. The exact solution for this case is
\begin{align*}
&{\vt r}(t) = {\vt r}_0 + \frac{t}{m} {\vt p}_0 \\
&{\vt p}(t) = {\vt p}_0 \\
&{\vt q}(t) = {\vt q}_0 \\
&{\vt \pi}(t) = {\vt \pi}_0
\end{align*}

@@

The Trotter-Suzuki splitting formula\cite{Trotter1959, Suzuki1976} can be used to devise a numerical solution for the rigid body motion, which still preserves the symplectic structure of Eq.~\ref{eq:EDO_system_alternative}. For a sum of two operators $i\!L_A$ and $i\!L_B$, such formula is
\[
e^{t(i\!L_A + i\!L_B)} \approx \left( e^{\frac{\Delta t}{2} i\!L_B} e^{\Delta t i\!L_A} e^{\frac{\Delta t}{2} i\!L_B} \right)^N + \mathcal{O}(\Delta t^2),
\]
where $\Delta t = t/N$ is the time step size and $N$ is the number of steps. We recall that each exponential operator above will act having as initial condition the result of the operator at its right-hand side. Splitting schemes with higher-order approximations have been used for rigid body dynamics,\cite{Omelyan2007, Omelyan2008, vanZon2008} but they result in methods which are more complex and often require expensive force-gradient calculations. Moreover, it is unclear whether their benefits would extend to non-Hamiltonian numerical integrators, as required for molecular dynamics in ensembles other than NVE.

@@

As recommended in Ref.~\onlinecite{Martyna1992}, the inertial parameters can be calculated from a characteristic time scale $t_d$ of the system by $Q_1 = L k_B T t_d^2$ and, for $j > 1$, $Q_j = k_B T t_d^2$. Analysis of the probability distribution generated by the equations of motion in Eq.~\ref{eq:nhc_system} can be done via the generalized phase-space analysis of Tuckerman and coworkers.\cite{Tuckerman2001} For this, conservation laws must be identified and driven variables must be eliminated. For a system without external forces, the extended energy $H$ and the vector quantity $\vt P e^{\eta_1} = \vt K$ are integrals of motion, while the center-of-mass position $\vt R$ can be considered as a driven variable due to the conservation of $\boldsymbol{\mathcal G}$.\cite{Tuckerman2001} Moreover, from the $8N$ individual equations implied in Eqs.~\ref{eq:nhc_q} and \ref{eq:nhc_pi}, $2N$ must be eliminated due to the constraints involving $\vt q_i$ and $\vt \pi_i$ for all $i$. By taking all these facts into account and applying the analysis of Ref.~\onlinecite{Tuckerman2001}, we can deduce that the metric determinant factor of the extended phase space is $\sqrt{g} = e^{6N \eta_1 + \sum_{j=2}^M \eta_j}$ and that the correct canonical distribution is attained if $L = 6N$.

\subsection{Integration Schemes for NVT Molecular Dynamics}

As done in the Hamiltonian case, the evolution of a phase-space function $f(\vt r, \vt p, \vt q, \vt \pi, \vt \eta,{\vt p}_\eta)$ is given by the relation $f(t) = e^{t i\!L}f_0$, but now in the context of the generalized Liouville theorem.\cite{Tuckerman_1999} This implies that we can employ the Trotter-Suzuki splitting to devise a strategy for integrating the equations of motion. First, we consider the usual separation of the non-Hamiltonian components of the equations of motion from the Hamiltonian components. In this case, the Liouville operator can be written as $i\!L = i\!L_\text{NHC} + i\!L_\text{NVE}$, where $i\!L_\text{NVE}$ was defined in Eq.~\ref{eq:full_operator}. It follows that the operator $i\!L_\text{NHC}$ is given by
\begin{equation}
\begin{split}
i\!L_\text{NHC} &= -\alpha_1 \sum_{i=1}^N \left( \tr{\vt p}_i \diff{}{\vt p_i} + \tr{\vt \pi}_i \diff{}{\vt \pi_i}\right) + \\
&+ \sum_{j=1}^M \alpha_j \diff{}{\eta_j} + \sum_{j=1}^M \left[ (G_j - \alpha_{j+1} p_{\eta_j}) \diff{}{p_{\eta_j}} \right].
\end{split}
\end{equation}

If the last summation was absent from the equation above, then the action of the propagator $e^{t i\!L_\text{NHC}}$ would be the solution of the equations $\dot{\vt p}_i = -\alpha_1 \vt p_i$ and $\dot{\vt \pi}_i = -\alpha_1 \vt \pi_i$ for all $i \in [1,N]$ and $\dot{\vt p}_i = \alpha_j$ for all $j \in [1,M]$, with all other variables remaining constant at their initial values. These equations can be solved altogether in analytical form, which results in
\begin{subequations}
\begin{align}
&\vt \pi_i = \vt \pi_i^0 \exp\left(-\alpha_1^0 t \right), \\
&\vt p_i = \vt p_i^0 \exp\left(-\alpha_1^0 t \right), \quad \text{and}\\
&\eta_j = \eta_j^0 + \alpha_j^0 t,
\end{align}
\end{subequations}
where the superscript $0$ stands for ``initial value''.

As pointed out by Martyna~\textit{et al}.,\cite{Martyna1996} the part of the operator given by $(G_j - \alpha_j p_{\eta_j}) \frac{\partial}{\partial p_{\eta_j}}$ can be evaluated analytically using the following equation:
\begin{equation}
\label{eq:p_eta}
\begin{split}
&p_{\eta_j} = p_{\eta_j}^0 + \left( G_j^0 - \alpha_{j+1}^0 p_{\eta_j}^0 \right) \phi(\alpha_{j+1}^0 t) t, \\
\end{split}
\end{equation}
where $\phi(x) = (1-e^{-x})/x$. For small absolute values of $x$, this function can be evaluated by truncating its Taylor series form $\phi(x) = \sum_{n=0}^\infty \frac{(-1)^n x^n}{(n+1)!}$. Four terms result in a sufficiently precise value if $\|x\| \leq 10^{-4}$. Considering the analytical solution above, the NHC operator is factorized as
\begin{equation}
\begin{split}
e^{\frac{\Delta t}{2} i\!L_\text{NHC}} &= \prod_{j=M}^1 e^{\frac{\Delta t}{4} (G_j - \alpha_{j+1} p_{\eta_j}) \frac{\partial}{\partial p_{\eta_j}}} \\
&\times e^{\frac{\Delta t}{2} \left[-\alpha_1 \sum\limits_{i=1}^N \left(\tr{\vt p}_i \frac{\partial}{\partial \vt p_i} + \tr{\vt \pi}_i \frac{\partial}{\partial \vt \pi_i}\right) + \sum\limits_{j=1}^M \alpha_j \frac{\partial}{\partial \eta_j} \right]} \\
&\times \prod_{j=1}^M e^{\frac{\Delta t}{4} (G_j - \alpha_{j+1} p_{\eta_j}) \frac{\partial}{\partial p_{\eta_j}}}.\\
\end{split}
\end{equation}

Retaining the NVE factorization scheme defined in Eq.~\ref{eq:trotter_splitting_NVE}, we write the full operator as

@@

>>>>

Analytical solutions exist for some simpler hypothetical cases related to Eq.~\ref{eq:EDO_system_alternative}. Let us analyze three of them, which will be useful later on to derive a complete approximate solution. In what follows, an index $0$ stands for ``initial value''. We start by considering a hypothetical Hamiltonian $\mathcal{H}_t = (1/2m) \tr{\vt p} \vt p$, whose resulting equations of motion are $\dot{\vt r} = {\vt p}/m$, $\dot{\vt p} = 0$, $\dot{\vt q} = 0$, and $\dot{\vt \pi} = 0$. The exact solution for this case is
\begin{align*}
&{\vt r}(t) = {\vt r}_0 + \frac{t}{m} {\vt p}_0 \\
&{\vt p}(t) = {\vt p}_0 \\
&{\vt q}(t) = {\vt q}_0 \\
&{\vt \pi}(t) = {\vt \pi}_0
\end{align*}

@@

The Trotter-Suzuki splitting formula\cite{Trotter1959, Suzuki1976} can be used to devise a numerical solution for the rigid body motion, which still preserves the symplectic structure of Eq.~\ref{eq:EDO_system_alternative}. For a sum of two operators $i\!L_A$ and $i\!L_B$, such formula is
\[
e^{t(i\!L_A + i\!L_B)} \approx \left( e^{\frac{\Delta t}{2} i\!L_B} e^{\Delta t i\!L_A} e^{\frac{\Delta t}{2} i\!L_B} \right)^\mathcal{N} + \mathcal{O}(\Delta t^2),
\]
where $\Delta t = t/\mathcal N$ is the time step size and $\mathcal N$ is the number of steps. We recall that each exponential operator above will act having as initial condition the result of the operator at its right-hand side. Splitting schemes with higher-order approximations have been used for rigid body dynamics,\cite{Omelyan2007, Omelyan2008, vanZon2008} but they result in methods which are more complex and often require expensive force-gradient calculations. Moreover, it is unclear whether their benefits would extend to non-Hamiltonian numerical integrators, as required for molecular dynamics in ensembles other than NVE.

@@

As recommended in Ref.~\onlinecite{Martyna1992}, the inertial parameters can be calculated from a characteristic time scale $t_d$ of the system by $Q_1 = L k_B T t_d^2$ and, for $j > 1$, $Q_j = k_B T t_d^2$. Analysis of the probability distribution generated by the equations of motion in Eq.~\ref{eq:nhc_system} can be done via the generalized phase-space analysis of Tuckerman and coworkers.\cite{Tuckerman2001} For this, conservation laws must be identified and driven variables must be eliminated. For a system without external forces, the extended energy $H$ and the vector quantity $\vt P e^{\eta_1} = \vt K$ are integrals of motion, while the center-of-mass position $\vt R$ can be considered as a driven variable due to the conservation of $\boldsymbol{\mathcal G}$.\cite{Tuckerman2001} Moreover, from the $8N$ individual equations implied in Eqs.~\ref{eq:nhc_q} and \ref{eq:nhc_pi}, $2N$ must be eliminated due to the constraints involving $\vt q_i$ and $\vt \pi_i$ for all $i$. By taking all these facts into account and applying the analysis of Ref.~\onlinecite{Tuckerman2001}, we can deduce that the metric determinant factor of the effective phase space is $\sqrt{g} = e^{(6N-2) \eta_1 + \sum_{j=2}^M \eta_j}$ and that the correct canonical distribution is attained if $L = 6N$.

\subsection{Integration Schemes for the Canonical Ensemble}

As done in the Hamiltonian case, the evolution of a phase-space function $f(\vt r, \vt p, \vt q, \vt \pi, \vt \eta,{\vt p}_\eta)$ is given by the relation $f(t) = e^{t i\!L}f_0$, but now in the context of the generalized Liouville theorem.\cite{Tuckerman_1999} This implies that we can employ the Trotter-Suzuki splitting to devise a strategy for integrating the equations of motion. First, we consider the usual separation of the non-Hamiltonian components of the equations of motion from the Hamiltonian components. In this case, the generalized Liouville operator can be written as $i\!L = i\!L_\text{NHC} + i\!L_\text{NVE}$, where $i\!L_\text{NVE}$ is defined as in Eq.~\ref{eq:full_operator}. Retaining the NVE factorization scheme defined in Eq.~\ref{eq:trotter_splitting_NVE}, we write the full operator as

@@

It follows from Eq.~\ref{eq:nhc_system} that
\begin{equation}
\label{eq:iL_NHC}
\begin{split}
i\!L_\text{NHC} &= -\alpha_1 \sum_{i=1}^N \left( \tr{\vt p}_i \diff{}{\vt p_i} + \tr{\vt \pi}_i \diff{}{\vt \pi_i}\right) + \\
&+ \sum_{j=1}^M \alpha_j \diff{}{\eta_j} + \sum_{j=1}^M \left[ (G_j - \alpha_{j+1} p_{\eta_j}) \diff{}{p_{\eta_j}} \right].
\end{split}
\end{equation}

If the last summation was absent from the equation above, then the action of the propagator $e^{t i\!L_\text{NHC}}$ would be the solution of the equations $\dot{\vt p}_i = -\alpha_1 \vt p_i$ and $\dot{\vt \pi}_i = -\alpha_1 \vt \pi_i$ for all $i \in [1,N]$ and $\dot{\vt p}_i = \alpha_j$ for all $j \in [1,M]$, with all other variables remaining constant at their initial values. These equations can be solved altogether in analytical form, which results in
\begin{subequations}
\begin{align}
&\vt p_i = \vt p_i^0 e^{-\alpha_1^0 t}, \\
&\vt \pi_i = \vt \pi_i^0 e^{-\alpha_1^0 t}, \quad \text{and}\\
&\eta_j = \eta_j^0 + \alpha_j^0 t.
\end{align}
\end{subequations}

On the other hand, if a single term of the last summation was the only term present in Eq.~\ref{eq:iL_NHC}, then $e^{t i\!L_\text{NHC}}$ would behave according to the corresponding equation $\dot{p}_{\eta_j} = G_j - \alpha_{j+1} p_{\eta_j}$, with all variables other than $p_{\eta_j}$ kept unchanged. The analytical solution of this single equation is
\begin{equation}
\label{eq:p_eta}
\begin{split}
&p_{\eta_j} = p_{\eta_j}^0 + \left( G_j^0 - \alpha_{j+1}^0 p_{\eta_j}^0 \right) \phi\left(\alpha_{j+1}^0 t\right) t, \\
\end{split}
\end{equation}
where $\phi(x) = (1-e^{-x})/x$. This is an alternative to the form found in Ref.~\onlinecite{Martyna1996}. We might note here that, in order to avoid numerical issues with $x \rightarrow 0$, we evaluate $\phi(x)$ as $\sum_{n=0}^4 \frac{(-1)^n x^n}{(n+1)!}$ whenever  $|x| \leq 10^{-4}$.

Considering the analytical solutions above, one can split the NHC operator into $2M+1$ factors by applying the Trotter-Suzuki formula recursively. The splitting we employ here is
\begin{equation}
\begin{split}
e^{\frac{\Delta t}{2} i\!L_\text{NHC}} &= \prod_{j=M}^1 e^{\frac{\Delta t}{4} (G_j - \alpha_{j+1} p_{\eta_j}) \frac{\partial}{\partial p_{\eta_j}}} \\
&\times e^{\frac{\Delta t}{2} \left[-\alpha_1 \sum\limits_{i=1}^N \left(\tr{\vt p}_i \frac{\partial}{\partial \vt p_i} + \tr{\vt \pi}_i \frac{\partial}{\partial \vt \pi_i}\right) + \sum\limits_{j=1}^M \alpha_j \frac{\partial}{\partial \eta_j} \right]} \\
&\times \prod_{j=1}^M e^{\frac{\Delta t}{4} (G_j - \alpha_{j+1} p_{\eta_j}) \frac{\partial}{\partial p_{\eta_j}}}.\\
\end{split}
\end{equation}
